Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-23 manuscript_rev1.tex
Rename manuscript

\usepackage{xr}
\externaldocument{si_rev1}

@@

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction and very low-level ($<$~250~m) cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subset our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog and very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

Fig.~\ref{fig:cyclone-stability}c, d show the geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section~\ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (50–65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are common elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilizing effect \cite{knight2024}. ICON is also substantially less stable than ERA5 across the whole region. In Section~\ref{sec:thermodynamic-profiles} we show that based on radiosonde observations, the bias is in ICON and not ERA5, and it is the result of underestimated temperature at heights corresponding to 700 hPa, as well as overestimated near-surface (2~m) air temperature, characterized by a higher frequency of occurrence in the 1–7°C range compared to observations at radiosonde launch locations (Fig.~\ref{fig:stats-hist-surf}a). This may be related to large-scale circulation in ICON or radiative transfer biases.

@@

We used the ALCF to derive cloud occurrence by height and the total cloud fraction from observations, ICON, ERA5, and MERRA-2. The results for all campaigns individually are shown in Fig.~\ref{fig:cloud-occurrence-panel}. As shown in this figure, the biases are relatively consistent across the campaigns and longitudes. In addition, we aggregated the campaigns by calculating the averages and percentiles of all individual profiles, presented in Fig.~\ref{fig:cloud-occurrence}. The analysis shows that the total cloud fraction is underestimated in ICON by about 10\% and in the reanalyses by about 20\%. When analyzed by height, ICON overestimates cloud occurrence below 1 km and underestimates it above; MERRA-2 underestimates cloud occurrence at all heights by up to 10\%, especially near the surface; and ERA5 simulates cloud occurrence relatively well above 1 km but strongly underestimates it near the surface. We note that fog or very low-level clouds are strongly underestimated in the reanalyses (fog and clouds are both included in the cloud occurrence). We conclude that the ICON results match the observations better than the reanalyses in this metric.

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10–30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the LCL distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air–sea moisture flux parameterization, or weaker stability in the models, which can cause more boundary mixing across heights and thus lower near-surface RH.

@@

Fig.~\ref{fig:stats-hist-si} shows the LWP and IWP distributions as histograms and their corresponding averages. The LWP and IWP are calculated from the mass of water in the column divided by the area of the column, i.e., not just the area of the cloudy portion of the column, as in some definitions. The available observational satellite reference for the LWP and IWP over high latitudes is unfortunately very uncertain due to a high solar zenith angle and the inability of passive visible and infrared retrievals to detect phase below the cloud top of mixed-phase clouds \cite{huang2006,greenwald2009,seethala2010,eliasson2011,duncan2018,khanal2020}, and this limits our comparison. The LWP distribution shows that all models overestimate cases with a near-zero LWP (Fig.~\ref{fig:stats-hist-si}a), which relates to the underestimated total cloud fraction. MERRA-2 shows quite overestimated high-LWP situations, which is most likely related to the too-few-too-bright problem of simulating lower total cloud fraction but clouds with a higher LWP to compensate. The IWP (Fig.~\ref{fig:stats-hist-si}b) is somewhat less important radiatively than LWP because of the typically larger and less numerous hydrometeors. Similarly to the LWP, the models overestimate situations with a near-zero IWP. ERA5 is otherwise simulating the IWP distribution well, but ICON and MERRA-2 underestimate the IWP. In the cloudy situations (Fig.~\ref{fig:stats-hist-si}c, d), it can be seen more distinctly that MERRA-2 overestimates moderate (0.05–0.15 kg m$^{-2}$) and high LWP (over 0.15 kg m$^{-2}$), and ERA5 and ICON underestimate moderate LWP. ICON also overestimates high LWP, resulting in overestimated average LWP.

@@

RH displays much larger biases. In all subsets, ICON is too humid in the first 1~km by about 5\%, but very accurate above, except for the 40–55°S zone and conditions of weak stability (Fig.~\ref{fig:potential-temperature}b, g), where it is too dry between about 1 and 3~km. Even though RH measured by radiosondes in the first 100~m is not very different between the observations and the models (Fig.~\ref{fig:potential-temperature}a), near-surface (2-m) RH at the radiosonde launch locations is much greater in the observations, most often close to 100\%, unlike in the models, where 85\% tends to be the most common (Fig.~\ref{fig:stats-hist-surf}b). This also explains why LCL is much more frequently located at the surface in the observations than in the models (Fig.~\ref{fig:cloud-occurrence}a). LCL is fully determined by near-surface temperature, near-surface RH, and surface pressure.

@@

Fig.~\ref{fig:rs-thetav-hur-cloud} shows $\theta_v$ and RH profiles for profiles containing fog, cloud at 500 m, and cloud at 1.5 km. These situations are characterized by particular cloud biases as identified in the lidar cloud occurrence analysis. The rationale is to examine $\theta_v$ and RH associated with these situations. Foggy situations are characterized by a rapid increase of $\theta_v$ with height and an observed average RH of about 90\% near the surface (Fig.~\ref{fig:rs-thetav-hur-cloud}a). In contrast, the models simulate higher RH in the first 100~m under foggy conditions by several percentage points. In situations with clouds occurring at 500 m, $\theta_v$ is relatively constant between the surface and 500 m (Fig.~\ref{fig:rs-thetav-hur-cloud}b), as expected for convectively driven clouds. The observed RH peaks at 500 m at about 90\%. The models, however, simulated higher RH between the surface and 500 m under these conditions. ICON and ERA5 show a stronger decrease of RH above this height than observations, and ERA5 shows more strongly stable stratification. Unlike the foggy and 500-m cloud situations, situations with clouds at 1.5 km do not have a flat $\theta_v$ with height. This indicates that, unlike the former, clouds at 1.5 km are not (or not as strongly) convectively driven. As expected, RH in these situations peaks at 1.5 km at about 85\% in observations. In the models, this peak is much less pronounced.

@@

Fig.~\ref{fig:rs-thetav-hist} shows histograms the same as the previous figure, but for $\theta_v$. They show a more complex picture, characterized by a central peak at about 0°C near the surface, increasing to about 5°C at 2 km (Fig.~\ref{fig:rs-thetav-hist}a). For cloudy bins, the central peak is generally more constant with height and even shows a minimum in $\theta_v$ at about 500 m (Fig.~\ref{fig:rs-thetav-hist}i). This is indicative of convection being associated with clouds at these heights, which results in flat $\theta_v$ profiles. In the reanalyses, in the first 200 m, values slightly above 0°C are associated with more clear bins than in observations, and values slightly below 0°C with fewer (Fig.~\ref{fig:rs-thetav-hist}g–h). Conversely, the opposite is true for cloudy bins (Fig.~\ref{fig:rs-thetav-hist}k–l). Situations with 0°C near-surface air temperature might occur predominantly when an open ocean surface keeps the near-surface air temperature close to 0°C under otherwise colder air mass conditions, such as under cold advection. ICON displays a notable bias above about 1 km, where the central peak is strongly underestimated (Fig.~\ref{fig:rs-thetav-hist}j). Instead, these heights and values of $\theta_v$ are more associated with clear bins (Fig.~\ref{fig:rs-thetav-hist}f). This might be related to the strong underestimation of cloud occurrence at these heights.

@@

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations, such as the ENSO in the observations and ICON. The interannual variability in cloud occurrence in ICON can be seen in Fig.~\ref{fig:cloud-occurrence-panel}, where each year in ICON is represented by a separate line. As could be expected, the interannual variability tends to be substantially smaller than the biases and thus is unlikely to have a strong impact on the main findings.

@@

We show that relative to ERA5, the distribution and strength of cyclonic activity over the SO is well represented in ICON, but it displays lower values of LTS. The latter is also manifested in the radiosonde profile comparison (Fig.~\ref{fig:rs-lts-hist}), showing that the $\theta_v$ profiles in ICON are less stable than in the observations. It is also manifested in near-surface air temperature, which is overestimated in the 1–7°C range at the radiosonde launch locations (Fig.~\ref{fig:stats-hist-surf}a). The underestimated LTS is linked to the overestimated cloud peak at 500 m in the lidar cloud occurrence comparison (Fig~\ref{fig:cloud-occurrence}f–g). It might also be interacting with the cloud inhomogeneity factor employed in ICON (Section~\ref{sec:icon}), resulting in lower cloud liquid water used in radiative calculations, hence decreased outgoing SW radiation. Based on the $\theta_v$ profile analysis, clouds at 500 m are predominantly convectively driven, and it is therefore expected that a model bias towards weak stability results in an increased cloud formation at this level. The underestimation of clouds above 1 km in ICON does not have a clear physical reason in our analysis and is likely partially or fully caused by stronger obscuration of the simulated lidar signal by the underlying and overestimated clouds in ICON at around 500 m.

The campaigns show remarkably similar biases in cloud occurrence by height in the lidar comparison (Fig.~\ref{fig:cloud-occurrence-panel}), which indicates that common underlying causes for the biases exist regardless of longitude and season. ICON underestimates the total cloud fraction by about 10\%, with an overestimation of clouds below 1~km and an underestimation of clouds above 1~km. The reanalyses underestimate the total cloud fraction by about 20\%. ERA5 overestimates clouds below 1~km but underestimates very low-level clouds and fog. ICON strongly overestimates the peak of cloud occurrence at about 500 m. This can be explained by the radiosonde comparison, showing that it is too moist at around this height (Fig.~\ref{fig:potential-temperature}a); has underestimated LTS (Fig.~\ref{fig:cyclone-stability} and \ref{fig:rs-lts-hist}), permitting shallow convection to this height; and has underestimated near-surface RH (Fig.~\ref{fig:stats-hist-surf}), resulting in higher LCL (Fig.~\ref{fig:cloud-occurrence}). Similar to our results for mid-level clouds, \citeA{cesana2022} showed that CMIP6 models also tend to underestimate cloud occurrence above 2~km over the SO, although their analysis in this case was limited to liquid clouds.

The inability of the models to simulate fog can be linked to various biases identified in our analysis. Near-surface RH is too low in the models (Fig.~\ref{fig:stats-hist-surf}), potentially due to low moisture flux from the surface and too effective boundary layer mixing. Near-surface temperature is also too high in ICON, and it can be expected that fog formation occurs in low near-surface temperature conditions when a warm and moist air mass is cooled by the surface to the saturation point. Fig.~\ref{fig:rs-thetav-hur-cloud} shows that fog occurs under highly stratified conditions. The underestimated LTS in ICON (and to a lesser extent in the reanalyses; Fig.~\ref{fig:rs-lts-hist}) indicates that the models are biased to weaker stability, thus having less favorable conditions for fog formation and persistence. The RH distribution in cloudy bins (Fig.~\ref{fig:rs-hur-hist}) also suggests that in observations, very low-level hydrometeors can occur under lower RH in observations than in the models. This could be due to high availability of cloud condensation nuclei (CCN) or ice nucleating particles (INPs) or due to hydrometeors and aerosols formed via sea spray under high swell and wind conditions. These parametrizations are likely very uncertain in the models in the SO due to the sparsity of reference data. \citeA{kawai2016} have shown that marine fog has some of the highest concentrations globally over the SO, and SO marine fog has a greater occurrence in winter. They conclude that marine fog is related to large-scale circulation and warm advection, and this is expected to change in a warming climate.

@@

The radiosonde observations indicate that the LCL is too high in ICON and reanalyses, which is probably responsible for the higher peak of clouds in the models and the lack of very low-level clouds and fog. Notably, ICON exhibits smaller internal variability in $\theta_v$ than the radiosonde observations. The analysis of the LWP and IWP (Fig.~\ref{fig:stats-hist-si}) shows that both phases are present in observations in about equal amounts. The models show diverse biases, the most pronounced being overestimation of high-LWP values in MERRA-2 and overestimation of cases with a near-zero LWP and IWP in all models. All models tend to compensate for the overestimated cases of a near-zero LWP with more high-LWP values to get a mean LWP that is either less (but close) to the observations (ERA5) or higher than the observations (ICON and MERRA-2). The IWP is underestimated in all of the models. In the case of ICON and MERRA-2, the mean IWP was underestimated and LWP overestimated, indicating that the models produce too much liquid and not enough ice phase. This is in contrast with previous findings of the lack of supercooled liquid over the SO in other models. If the liquid phase is overestimated relative to the ice phase, one would expect underestimated cloud SW reflectivity due to a larger number of smaller hydrometeors for the same amount of water. Cloudy areas would then appear brighter in the SW spectrum. This can contribute to the too few, too bright bias, i.e., the overestimated brightness of cloudy areas compensates for the lower total cloud fraction in the models. As mentioned in Section~\ref{sec:rad-lwp-iwp}, the LWP and IWP are, however, affected by the high uncertainty of the satellite retrievals.

>>>>

@@

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction and very low-level cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON and two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subset our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog and very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

Fig.~\ref{fig:cyclone-stability}c, d show the geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section~\ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (50–65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are common elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilizing effect \cite{knight2024}. ICON is also substantially less stable than ERA5 across the whole region. In Section~\ref{sec:thermodynamic-profiles} we show that based on radiosonde observations, the bias is in ICON and not ERA5, and it is the result of underestimated temperature at heights corresponding to 700 hPa, as well as overestimated near-surface (2~m) air temperature, characterized by a higher frequency of occurrence in the 1–7°C range compared to observations at radiosonde launch locations (Fig.~S1a). This may be related to large-scale circulation in ICON or radiative transfer biases.

@@

We used the ALCF to derive cloud occurrence by height and the total cloud fraction from observations, ICON, ERA5, and MERRA-2. The results for all campaigns individually are shown in Fig.~S2. As shown in this figure, the biases are relatively consistent across the campaigns and longitudes. In addition, we aggregated the campaigns by calculating the averages and percentiles of all individual profiles, presented in Fig.~\ref{fig:cloud-occurrence}. The analysis shows that the total cloud fraction is underestimated in ICON by about 10\% and in the reanalyses by about 20\%. When analyzed by height, ICON overestimates cloud occurrence below 1 km and underestimates it above; MERRA-2 underestimates cloud occurrence at all heights by up to 10\%, especially near the surface; and ERA5 simulates cloud occurrence relatively well above 1 km but strongly underestimates it near the surface. We note that fog or very low-level clouds are strongly underestimated in the reanalyses (fog and clouds are both included in the cloud occurrence). We conclude that the ICON results match the observations better than the reanalyses in this metric.

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10–30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the LCL distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~S1b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air–sea moisture flux parameterization, or weaker stability in the models, which can cause more boundary mixing across heights and thus lower near-surface RH.

@@

Fig.~S3 shows the LWP and IWP distributions as histograms and their corresponding averages. The LWP and IWP are calculated from the mass of water in the column divided by the area of the column, i.e., not just the area of the cloudy portion of the column, as in some definitions. The available observational satellite reference for the LWP and IWP over high latitudes is unfortunately very uncertain due to a high solar zenith angle and the inability of passive visible and infrared retrievals to detect phase below the cloud top of mixed-phase clouds \cite{huang2006,greenwald2009,seethala2010,eliasson2011,duncan2018,khanal2020}, and this limits our comparison. The LWP distribution shows that all models overestimate cases with a near-zero LWP (Fig.~S3a), which relates to the underestimated total cloud fraction. MERRA-2 shows quite overestimated high-LWP situations, which is most likely related to the too-few-too-bright problem of simulating lower total cloud fraction but clouds with a higher LWP to compensate. The IWP (Fig.~S3b) is somewhat less important radiatively than LWP because of the typically larger and less numerous hydrometeors. Similarly to the LWP, the models overestimate situations with a near-zero IWP. ERA5 is otherwise simulating the IWP distribution well, but ICON and MERRA-2 underestimate the IWP. In the cloudy situations (Fig.~S3c, d), it can be seen more distinctly that MERRA-2 overestimates moderate (0.05–0.15 kg m$^{-2}$) and high LWP (over 0.15 kg m$^{-2}$), and ERA5 and ICON underestimate moderate LWP. ICON also overestimates high LWP, resulting in overestimated average LWP.

@@

RH displays much larger biases. In all subsets, ICON is too humid in the first 1~km by about 5\%, but very accurate above, except for the 40–55°S zone and conditions of weak stability (Fig.~\ref{fig:potential-temperature}b, g), where it is too dry between about 1 and 3~km. Even though RH measured by radiosondes in the first 100~m is not very different between the observations and the models (Fig.~\ref{fig:potential-temperature}a), near-surface (2-m) RH at the radiosonde launch locations is much greater in the observations, most often close to 100\%, unlike in the models, where 85\% tends to be the most common (Fig.~S1b). This also explains why LCL is much more frequently located at the surface in the observations than in the models (Fig.~\ref{fig:cloud-occurrence}a). LCL is fully determined by near-surface temperature, near-surface RH, and surface pressure.

@@

Fig.~S4 shows $\theta_v$ and RH profiles for profiles containing fog, cloud at 500 m, and cloud at 1.5 km. These situations are characterized by particular cloud biases as identified in the lidar cloud occurrence analysis. The rationale is to examine $\theta_v$ and RH associated with these situations. Foggy situations are characterized by a rapid increase of $\theta_v$ with height and an observed average RH of about 90\% near the surface (Fig.~S4a). In contrast, the models simulate higher RH in the first 100~m under foggy conditions by several percentage points. In situations with clouds occurring at 500 m, $\theta_v$ is relatively constant between the surface and 500 m (Fig.~S4b), as expected for convectively driven clouds. The observed RH peaks at 500 m at about 90\%. The models, however, simulated higher RH between the surface and 500 m under these conditions. ICON and ERA5 show a stronger decrease of RH above this height than observations, and ERA5 shows more strongly stable stratification. Unlike the foggy and 500-m cloud situations, situations with clouds at 1.5 km do not have a flat $\theta_v$ with height. This indicates that, unlike the former, clouds at 1.5 km are not (or not as strongly) convectively driven. As expected, RH in these situations peaks at 1.5 km at about 85\% in observations. In the models, this peak is much less pronounced.

@@

Fig.~S5 shows histograms the same as the previous figure, but for $\theta_v$. They show a more complex picture, characterized by a central peak at about 0°C near the surface, increasing to about 5°C at 2 km (Fig.~S5a). For cloudy bins, the central peak is generally more constant with height and even shows a minimum in $\theta_v$ at about 500 m (Fig.~S5i). This is indicative of convection being associated with clouds at these heights, which results in flat $\theta_v$ profiles. In the reanalyses, in the first 200 m, values slightly above 0°C are associated with more clear bins than in observations, and values slightly below 0°C with fewer (Fig.~S5g–h). Conversely, the opposite is true for cloudy bins (Fig.~S5k–l). Situations with 0°C near-surface air temperature might occur predominantly when an open ocean surface keeps the near-surface air temperature close to 0°C under otherwise colder air mass conditions, such as under cold advection. ICON displays a notable bias above about 1 km, where the central peak is strongly underestimated (Fig.~S5j). Instead, these heights and values of $\theta_v$ are more associated with clear bins (Fig.~S5f). This might be related to the strong underestimation of cloud occurrence at these heights.

@@

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations, such as the ENSO in the observations and ICON. The interannual variability in cloud occurrence in ICON can be seen in Fig.~S2, where each year in ICON is represented by a separate line. As could be expected, the interannual variability tends to be substantially smaller than the biases and thus is unlikely to have a strong impact on the main findings.

@@

We show that relative to ERA5, the distribution and strength of cyclonic activity over the SO is well represented in ICON, but it displays lower values of LTS. The latter is also manifested in the radiosonde profile comparison (Fig.~\ref{fig:rs-lts-hist}), showing that the $\theta_v$ profiles in ICON are less stable than in the observations. It is also manifested in near-surface air temperature, which is overestimated in the 1–7°C range at the radiosonde launch locations (Fig.~S1a). The underestimated LTS is linked to the overestimated cloud peak at 500 m in the lidar cloud occurrence comparison (Fig~\ref{fig:cloud-occurrence}f–g). It might also be interacting with the cloud inhomogeneity factor employed in ICON (Section~\ref{sec:icon}), resulting in lower cloud liquid water used in radiative calculations, hence decreased outgoing SW radiation. Based on the $\theta_v$ profile analysis, clouds at 500 m are predominantly convectively driven, and it is therefore expected that a model bias towards weak stability results in an increased cloud formation at this level. The underestimation of clouds above 1 km in ICON does not have a clear physical reason in our analysis and is likely partially or fully caused by stronger obscuration of the simulated lidar signal by the underlying and overestimated clouds in ICON at around 500 m.

The campaigns show remarkably similar biases in cloud occurrence by height in the lidar comparison (Fig.~S2), which indicates that common underlying causes for the biases exist regardless of longitude and season. ICON underestimates the total cloud fraction by about 10\%, with an overestimation of clouds below 1~km and an underestimation of clouds above 1~km. The reanalyses underestimate the total cloud fraction by about 20\%. ERA5 overestimates clouds below 1~km but underestimates very low-level clouds and fog. ICON strongly overestimates the peak of cloud occurrence at about 500 m. This can be explained by the radiosonde comparison, showing that it is too moist at around this height (Fig.~\ref{fig:potential-temperature}a); has underestimated LTS (Fig.~\ref{fig:cyclone-stability} and \ref{fig:rs-lts-hist}), permitting shallow convection to this height; and has underestimated near-surface RH (Fig.~S1), resulting in higher LCL (Fig.~\ref{fig:cloud-occurrence}). Similar to our results for mid-level clouds, \citeA{cesana2022} showed that CMIP6 models also tend to underestimate cloud occurrence above 2~km over the SO, although their analysis in this case was limited to liquid clouds.

The inability of the models to simulate fog can be linked to various biases identified in our analysis. Near-surface RH is too low in the models (Fig.~S1), potentially due to low moisture flux from the surface and too effective boundary layer mixing. Near-surface temperature is also too high in ICON, and it can be expected that fog formation occurs in low near-surface temperature conditions when a warm and moist air mass is cooled by the surface to the saturation point. Fig.~S4 shows that fog occurs under highly stratified conditions. The underestimated LTS in ICON (and to a lesser extent in the reanalyses; Fig.~\ref{fig:rs-lts-hist}) indicates that the models are biased to weaker stability, thus having less favorable conditions for fog formation and persistence. The RH distribution in cloudy bins (Fig.~\ref{fig:rs-hur-hist}) also suggests that in observations, very low-level hydrometeors can occur under lower RH in observations than in the models. This could be due to high availability of cloud condensation nuclei (CCN) or ice nucleating particles (INPs) or due to hydrometeors and aerosols formed via sea spray under high swell and wind conditions. These parametrizations are likely very uncertain in the models in the SO due to the sparsity of reference data. \citeA{kawai2016} have shown that marine fog has some of the highest concentrations globally over the SO, and SO marine fog has a greater occurrence in winter. They conclude that marine fog is related to large-scale circulation and warm advection, and this is expected to change in a warming climate.

@@

The radiosonde observations indicate that the LCL is too high in ICON and reanalyses, which is probably responsible for the higher peak of clouds in the models and the lack of very low-level clouds and fog. Notably, ICON exhibits smaller internal variability in $\theta_v$ than the radiosonde observations. The analysis of the LWP and IWP (Fig.~S3) shows that both phases are present in observations in about equal amounts. The models show diverse biases, the most pronounced being overestimation of high-LWP values in MERRA-2 and overestimation of cases with a near-zero LWP and IWP in all models. All models tend to compensate for the overestimated cases of a near-zero LWP with more high-LWP values to get a mean LWP that is either less (but close) to the observations (ERA5) or higher than the observations (ICON and MERRA-2). The IWP is underestimated in all of the models. In the case of ICON and MERRA-2, the mean IWP was underestimated and LWP overestimated, indicating that the models produce too much liquid and not enough ice phase. This is in contrast with previous findings of the lack of supercooled liquid over the SO in other models. If the liquid phase is overestimated relative to the ice phase, one would expect underestimated cloud SW reflectivity due to a larger number of smaller hydrometeors for the same amount of water. Cloudy areas would then appear brighter in the SW spectrum. This can contribute to the too few, too bright bias, i.e., the overestimated brightness of cloudy areas compensates for the lower total cloud fraction in the models. As mentioned in Section~\ref{sec:rad-lwp-iwp}, the LWP and IWP are, however, affected by the high uncertainty of the satellite retrievals.
